---
title: "Proj1"
author: "Thibeaux"
date: "2023-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Libraries
```{r}
library(GGally)
library(boot)
library(ggplot2)
library(tidyverse)
library(lmboot)
library(lattice)
library(caret)
library(naniar)
library(utils)
library(stats)
library(corrplot)
library(ISLR)
library(car)
library(olsrr)
library(data.table)
library(GGally)
```

## Reading in Dataset
```{r}
life = read.csv('https://github.com/athibeaux/MSDS-DS6372/raw/main/Life_Expectancy.csv')
```

```{r}
ggplot(data = life) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))
```
Upon looking at the graph of the original data set, it appears that there needs to be a log transformation on the X or the GDP as we are interested in seeing the relation between Life Expenctancy and GDP. 

## Checking Data Types
```{r}
str(life)
vis_miss(life)
dim(life)
#View(life)
#sum(is.na(life))
```

# Imputing using Median
```{r}
#GDP 15% [17]  keep GDP to have it Imputed even if quite high percentage, assuming it is crucial to predicting Life.expectancy as richer countries have better health access/Medicine and tech. The numbers appear to be GDP per capita which helps as it addresses GDP/Population. GDP per Capita and Population would be too closely related and prob attribute to covariance.  
#Adjusting text angle to vis_miss
imputeMedian= preProcess(life[,-c(1:4,9)],method="medianImpute") #predictors 1:4, 9 and response is 4
cleandataMedian = predict(imputeMedian,newdata=life)
dim(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#Literature says that over 10% missing data can contribute to bias 
#HepatitsB [9] at 19% , Population 22% [18]. 
#Removing columns 9 and 18
cleandataMedian = cleandataMedian[,-c(18,9)]
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#removing last NA
cleandataMedian = na.omit(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

```
Creating Variable of Logged GDP Attribute
```{r}
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))

#Converting GDP to Log 
cleandataMedian$logGDP = log(cleandataMedian$GDP)
#converting Life.expectancy to log 
cleandataMedian$logLife.expectancy = log(cleandataMedian$Life.expectancy)

#Log transformation on GDP 
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = logGDP, y = logLife.expectancy))

```
Splitting the data
```{r}
set.seed(1234)
trainIndex<-createDataPartition(cleandataMedian$Life.expectancy,p=.8,list=F)  #p: proportion of data in train

training<-cleandataMedian[trainIndex,]
validate<-cleandataMedian[-trainIndex,]
```

Before and after log transforming GDP, with cleandataMedian:
```{r}
ggplot(data = training) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))

#Log transformation on GDP 
ggplot(data = training) + geom_point(mapping = aes(x = log(GDP), y = Life.expectancy))

```
# EDA

## World Map, Color plotting 
```{r}
#Creating the World 
#library(ggplot2)
#library(tidyverse)
#library(ggthemes)

#world_map = map_data("world") %>% filter(! long > 180)

#countries = world_map %>% distinct(region) %>% rowid_to_column()

#countries %>% ggplot(aes(fill = rowid, map_id = region)) + geom_map(map = world_map) + expand_limits(x = world_map$long, y = world_map$lat) + coord_map("moll") +theme_map()


```

## Color world plotting with full data
```{r}
library(ggplot2)
library(tidyverse)

#rename training for the map
dataforcolmap = training
#but first renaming column Country in dataforcolmap 
colnames(dataforcolmap)[1] = "region"



#Renaming United States of America and Boliva (Republic...) with USA and Boliva and other countries as follows 
dataforcolmap$region[dataforcolmap$region == "United States of America"] = "USA"
dataforcolmap$region[dataforcolmap$region == "Bolivia (Plurinational State of)"] = "Bolivia"
dataforcolmap$region[dataforcolmap$region == "Venezuela (Bolivarian Republic of)"] = "Venezuela"
dataforcolmap$region[dataforcolmap$region == "Republic of Korea"] = "South Korea"
dataforcolmap$region[dataforcolmap$region == "The former Yugoslav republic of Macedonia"] = "North Macedonia"
dataforcolmap$region[dataforcolmap$region == "Republic of Moldova"] = "Moldova"
dataforcolmap$region[dataforcolmap$region == "Russian Federation"] = "Russia"
dataforcolmap$region[dataforcolmap$region == "Micronesia (Federated States of)"] = "Micronesia"
dataforcolmap$region[dataforcolmap$region == "Lao People's Democratic Republic"] = "Laos"
dataforcolmap$region[dataforcolmap$region == "Iran (Islamic Republic of)"] = "Iran"
dataforcolmap$region[dataforcolmap$region == "Democratic People's Republic of Korea"] = "North Korea"


view(dataforcolmap)


#getting map data for plotting 
mapdata = map_data("world")
#view(mapdata)

#joining map data with dataforcolmap 
mapdata = left_join(mapdata,dataforcolmap, by = "region")
#view(mapdata)

#filtering out NAs for life expectancy , status, Income.composition.of.resources
#Life Exp
mapdata1 = mapdata %>% filter(!is.na(mapdata$Life.expectancy))

#Status 
mapdata2 = mapdata %>% filter(!is.na(mapdata$Status))

#Income
mapdata3 = mapdata %>% filter(!is.na(mapdata$Income.composition.of.resources))


#mapping mapdata1 for Life Exp 

map1 = ggplot(mapdata1, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Life.expectancy), color = "black")+ theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Life Expectancy per Country") + scale_fill_gradient(low = "red", high = "yellow") + guides(fill=guide_legend(title="Life Expectancy"))
map1



#mapping mapdata2 for Status 

mapStatus = ggplot(mapdata2, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Status, col = "orange"), color = "black") + theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Country's Status: Developed v. Developing") 
mapStatus


#mapping mapdata3 for Income Composition of Resources 

mapIncome = ggplot(mapdata3, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Income.composition.of.resources), color = "black")+ theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Income Composition of Resources per Country") + scale_fill_gradient(low = "red", high = "yellow")+ guides(fill=guide_legend(title="Income Composition of Resources"))
mapIncome

```

## Correlation Matrix:
```{r}
cor <- cor(training[,c(4,5:20)])
corrplot(cor, method = "square",  tl.srt = 50, tl.col = "black", tl.cex = 0.6, title = "Correlation of Variables", mar=c(0,0,1,0))
```
GGPairs:
```{r}
#commented out for knitting
#ggpairs(training[,4:20])
```
# Multivariable Plots on Full Data
```{r}
# Libraries Used: ISLR, data.table, GGally, ggplot2

ggpairs(training[,4:8], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,9:12)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,13:16)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,17:20)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))
#ggpairs(cleandataMedian[,5:22], upper = list(continuous = wrap("cor", size = 4.75, align_percent = 1)))
#ggscatmat(cleandataMedian, columns = 4:10)
```

Full Model
```{r}
KitchenSink <- lm(Life.expectancy~Year + Status + Adult.Mortality +
                  infant.deaths + Alcohol + percentage.expenditure +
                  Measles + BMI + under.five.deaths + Polio + 
                  Total.expenditure + Diphtheria + HIV.AIDS + log(GDP) + 
                    thinness..1.19.years + thinness.5.9.years +
                    Income.composition.of.resources + Schooling, training)
plot(KitchenSink)
vif(KitchenSink)

full.model <- lm(Life.expectancy~ Status + Alcohol + percentage.expenditure + 
                   Measles + BMI + under.five.deaths + Polio + 
                   Total.expenditure + Diphtheria + HIV.AIDS + log(GDP) + 
                   thinness..1.19.years +
                   Income.composition.of.resources + Schooling, training)

summary(full.model)
plot(full.model)
vif(full.model)
```
# Residuals Custom model 
```{r}
set.seed(2345)
eightVar = lm(Life.expectancy~HIV.AIDS+Schooling+Alcohol+BMI+Polio+Diphtheria+logGDP+thinness..1.19.years+Income.composition.of.resources, data = training)
summary(eightVar)
confint(eightVar)

#Visuals for Residuals
plot(eightVar)

```

#Best final MLR 
```{r}
set.seed(2323)
lessvar = lm(Life.expectancy~HIV.AIDS+Schooling+BMI+Diphtheria+logGDP+thinness..1.19.years+Income.composition.of.resources, data = training)
summary(lessvar)
confint(lessvar)

#Visuals for Residuals
plot(lessvar)
```

Feature Selection Tools: Penalized Regression
```{r Penalized Regression}

# Penalized Regression

#Setting kfold parameters
fitControl<-trainControl(method="repeatedcv",number=5,repeats=1) 

#Fitting glmnet
set.seed(1234)
glmnet.fit<-train(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling,
                  data=training,
                  method="glmnet",
                  trControl=fitControl
)

glmnet.fit
plot(glmnet.fit)


#Investigating coefficients
opt.pen<-glmnet.fit$finalModel$lambdaOpt #penalty term
coef(glmnet.fit$finalModel,opt.pen)


glmnet.fit.model <-lm(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years +
              Income.composition.of.resources + Schooling,
          data=training)

plot(glmnet.fit.model)

#Lets force a LASSO model and add complexity
set.seed(1234)
glmnet.fit2<-train(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + poly(Polio,2) + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling,
                  data=training,
                  method="glmnet",
                  trControl=fitControl,
                  tuneGrid=expand.grid(data.frame(alpha=1,lambda=seq(0,.05,.001)))
)

glmnet.fit2
plot(glmnet.fit2)

opt.pen<-glmnet.fit2$finalModel$lambdaOpt   #penalty term
coef(glmnet.fit2$finalModel,opt.pen)

# Different way to do GLMNET

x=model.matrix(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling,
               training)[,-1]
y=log(training$Life.expectancy)

library(glmnet)
set.seed(1234)

grid=10^seq(10,-2, length =100)
lasso.mod=glmnet(x,y,alpha=1, lambda =grid)

cv.out=cv.glmnet(x,y,alpha=1)
plot(cv.out)
bestlambda<-cv.out$lambda.1se
coef(cv.out,s=bestlambda)
```
Feature Selection Tools: Forward Selection
```{r Forward Selection with olsrr}
fwd.selection = lm(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling, data = training)

# Forward
ols_step_forward_p(fwd.selection, penter = 0.05, details = TRUE)

# Forward Selection Chosen Model

fwd.select.model <- lm(Life.expectancy~ Income.composition.of.resources + Schooling +
                         HIV.AIDS + Diphtheria +
                         BMI + log(GDP) + Polio +
                         percentage.expenditure + thinness..1.19.years + Total.expenditure, training)

plot(fwd.select.model)
```

```{r Forward Selection with Leaps}
library(leaps)
reg.fwd=regsubsets(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling,
                   data=training,method="forward",nvmax=15)

summary(reg.fwd)$adjr2
summary(reg.fwd)$rss
summary(reg.fwd)$bic


par(mfrow=c(1,3))
bics<-summary(reg.fwd)$bic
plot(1:15,bics,type="l",ylab="BIC",xlab="# of predictors")
index<-which(bics==min(bics))
points(index,bics[index],col="red",pch=10)

adjr2<-summary(reg.fwd)$adjr2
plot(1:15,adjr2,type="l",ylab="Adjusted R-squared",xlab="# of predictors")
index<-which(adjr2==max(adjr2))
points(index,adjr2[index],col="red",pch=10)

rss<-summary(reg.fwd)$rss
plot(1:15,rss,type="l",ylab="train RSS",xlab="# of predictors")
index<-which(rss==min(rss))
points(index,rss[index],col="red",pch=10)


```
Backward Selection
```{r Backward Selection with olsrr}
bck.selection = lm(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling, data = training)

# Backward
ols_step_backward_p(bck.selection, prem = 0.05, details = TRUE)

# Backward Selection Chosen Model
bck.select.model <- lm(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + 
              Income.composition.of.resources + Schooling, data = training)

plot(bck.select.model)
```

Stepwise Selection
```{r Stepwise Selection with olsrr}
sw.selection = lm(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling, data = training)
summary(sw.selection)

# Stepwise
ols_step_both_p(sw.selection, prem = 0.05, pent = 0.05, details = FALSE)

# Stepwise Selection Chosen Model
sw.select.model = lm(Life.expectancy~ percentage.expenditure + 
                       BMI + Polio + Diphtheria + HIV.AIDS + 
                       log(GDP) + thinness..1.19.years +
                       Income.composition.of.resources + Schooling, data = training)

plot(sw.select.model)
```

Forward Selection Validation
```{r Train Test Data}

fwd.train=regsubsets(Life.expectancy~infant.deaths + Alcohol + 
              percentage.expenditure + Measles + BMI + 
              under.five.deaths + Polio + Total.expenditure + 
              Diphtheria + HIV.AIDS + log(GDP) + 
              thinness..1.19.years + thinness.5.9.years +
              Income.composition.of.resources + Schooling,
                    data=training,method="forward",nvmax=15)

#Creating a prediction function 
predict.regsubsets =function (object , newdata ,id ,...){
  form=as.formula (object$call [[2]])
  mat=model.matrix(form ,newdata )
  coefi=coef(object ,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}

valMSE<-c()
#note my index, i, is to 15 since that is how many predictors I went up to during fwd selection
for (i in 1:15){
  predictions<-predict.regsubsets(object=fwd.train,newdata=validate,id=i) 
  valMSE[i]<-mean((validate$Life.expectancy-predictions)^2)
}
par(mfrow=c(1,1))
plot(1:15,sqrt(valMSE),type="l",xlab="# of predictors",
     ylab="test vs train RMSE")
index<-which(valMSE==min(valMSE))
points(index,sqrt(valMSE[index]),col="red",pch=10)

trainMSE<-summary(fwd.train)$rss/nrow(training)
lines(1:15,sqrt(trainMSE),lty=3,col="blue")  
```
# Forward, Backward, Stepwise Selection with AIC
```{r}
library(MASS)
#Full Model 
set.seed(1246)
fitFull = lm(Life.expectancy~Adult.Mortality + 
     infant.deaths + Alcohol + percentage.expenditure + Measles + 
     BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling + logGDP,training)

stepup = stepAIC(fitFull, direction = "forward", steps = 2000)

stepdown= stepAIC(fitFull, direction = "backward", steps = 2000)

stepboth = stepAIC(fitFull, direction = "both", steps = 2000)

#summary for each model 
up = summary(stepup)
up
down = summary(stepdown)
down
both = summary(stepboth)
both

fitFull

olsrr::ols_step_forward_aic(fitFull)
```
Adding Complexity to the Model
```{r Transformation Exploration}
# Transformations using Polio as a test

transformations <- training

# Original
training %>% ggplot(aes(x=Polio, y = Life.expectancy)) + 
           geom_point() + geom_smooth () + ylab("Life Expectancy in Age")

# Transforming with Polynomial using a ^2
transformations$polio2 <- transformations$Polio^2
transformations %>% ggplot(aes(x=polio2, y=Life.expectancy)) +
geom_point() + geom_smooth()

# Transforming using Log
transformations$logPolio <- log(transformations$Polio)

transformations %>% ggplot(aes(x=logPolio, y=Life.expectancy)) +
geom_point() + geom_smooth()

# Testing if it's the same as logging inside the plot
transformations %>% ggplot(aes(x=log(Polio), y=Life.expectancy)) +
geom_point() + geom_smooth()
```
Creating plot functions
```{r Creating Plot Functions}

# Original Data plot
plot_original <- function(var1) {
  return(training %>% ggplot(aes(x=var1, y = Life.expectancy)) + 
           geom_point() + geom_smooth () + ylab("Life Expectancy in Age")
)
}

plot_original_interaction <- function(var1) {
  return(training %>% ggplot(aes(x=var1, 
                                        y = Life.expectancy, 
                                        color=Status)) + 
           geom_point() + geom_smooth () + ylab("Life Expectancy in Age")
)
}

plot_original(training$Polio)+ 
  xlab("Polio Immunization Coverage among 1-year-olds")

# Creating function to plot variable as polynomials 

plot_poly <- function(var1) {
  return(training %>% ggplot(aes(x=var1^2, 
                                      y = Life.expectancy,
                                      color=Status)) + 
         geom_point() + geom_smooth () + ylab("Life Expectancy in Age")
)
}

plot_poly(training$Polio)+ 
  xlab("Polio Immunization Coverage among 1-year-olds")

# Creating function to plot variables as logged
plot_log <- function(var1) {
  return(training %>% ggplot(aes(x=log(var1), 
                                      y = Life.expectancy, 
                                      color=Status)) + 
         geom_point() + geom_smooth () + ylab("Life Expectancy in Age")
)
}

plot_log(training$Polio) + 
  xlab("Polio Immunization Coverage among 1-year-olds")
```
Putting each Variable into the Function
```{r Variables into Function}
# Infant Deaths Variable

plot_original(training$infant.deaths) + xlab("Infant Deaths per 1000")
plot_original_interaction(training$infant.deaths) + xlab("Infant Deaths per 1000")
plot_poly(training$infant.deaths) + xlab("Infant Deaths per 1000")
plot_log(training$infant.deaths) + xlab("Infant Deaths per 1000")

# Alcohol Variable

plot_original(training$Alcohol) + 
  xlab("Alcohol Consumption per capita in Litres of pure alcohol")
plot_original_interaction(training$Alcohol) + 
  xlab("Alcohol Consumption per capita in Litres of pure alcohol")
plot_poly(training$Alcohol) + 
  xlab("Alcohol Consumption per capita in Litres of pure alcohol")
plot_log(training$Alcohol) + 
  xlab("Alcohol Consumption per capita in Litres of pure alcohol")

# Percentage of Expenditures Variable

plot_original(training$percentage.expenditure) + 
  xlab("Expenditure on health as a percentage of Gross Domestic Product per capita")
plot_original_interaction(training$percentage.expenditure) + 
  xlab("Expenditure on health as a percentage of Gross Domestic Product per capita")
plot_poly(training$percentage.expenditure) + 
  xlab("Expenditure on health as a percentage of Gross Domestic Product per capita")
plot_log(training$percentage.expenditure) + 
  xlab("Expenditure on health as a percentage of Gross Domestic Product per capita")

# Measles Variable

plot_original(training$Measles) + xlab("Reported Measles cases per 1000")
plot_original_interaction(training$Measles) + xlab("Reported Measles cases per 1000")
plot_poly(training$Measles)+ xlab("Reported Measles cases per 1000")
plot_log(training$Measles)+ xlab("Reported Measles cases per 1000")

# BMI Variable

plot_original(training$BMI) + xlab("Average Body Mass Index of population")
plot_original_interaction(training$BMI) + xlab("Average Body Mass Index of population")
plot_poly(training$BMI)+ xlab("Average Body Mass Index of population")
plot_log(training$BMI)+ xlab("Average Body Mass Index of population")

# Under Five Deaths Variable

plot_original(training$under.five.deaths) + 
  xlab("Under age 5 deaths per 1000")
plot_original_interaction(training$under.five.deaths) + 
  xlab("Under age 5 deaths per 1000")
plot_poly(training$under.five.deaths)+ 
  xlab("Under age 5 deaths per 1000")
plot_log(training$under.five.deaths)+ 
  xlab("Under age 5 deaths per 1000")

# Total Expenditures Variable

plot_original(training$Total.expenditure) + 
  xlab("Percentage of total government expenditures on health")
plot_original_interaction(training$Total.expenditure) + 
  xlab("Percentage of total government expenditures on health")
plot_poly(training$Total.expenditure)+ 
  xlab("Percentage of total government expenditures on health")
plot_log(training$Total.expenditure)+ 
  xlab("Percentage of total government expenditures on health")

# Diphtheria Variable

plot_original(training$Diphtheria) + 
  xlab("DPT3 Immunization Coverage among 1-year-olds")
plot_original_interaction(training$Diphtheria) + 
  xlab("DPT3 Immunization Coverage among 1-year-olds")
plot_poly(training$Diphtheria)+ 
  xlab("DPT3 Immunization Coverage among 1-year-olds")
plot_log(training$Diphtheria)+ 
  xlab("DPT3 Immunization Coverage among 1-year-olds")

# HIV AIDS Variable

plot_original(training$HIV.AIDS) + xlab ("HIV/AIDS Deaths per 1000 live births")
plot_original_interaction(training$HIV.AIDS) + xlab ("HIV/AIDS Deaths per 1000 live births")
plot_poly(training$HIV.AIDS)+ xlab ("HIV/AIDS Deaths per 1000 live births")
plot_log(training$HIV.AIDS)+ xlab ("HIV/AIDS Deaths per 1000 live births")

# GDP Variable

plot_original(training$GDP) + xlab("Gross Domestic Product per capita in USD")
plot_original_interaction(training$GDP) + xlab("Gross Domestic Product per capita in USD")
plot_poly(training$GDP)+ xlab("Gross Domestic Product per capita in USD")
plot_log(training$GDP)+ xlab("Gross Domestic Product per capita in USD")

# Thinness 10-19 Years Variable

plot_original(training$thinness..1.19.years) + 
  xlab("Prevalence of thinness among children ages 10 to 19")
plot_original_interaction(training$thinness..1.19.years) + 
  xlab("Prevalence of thinness among children ages 10 to 19")
plot_poly(training$thinness..1.19.years)+ 
  xlab("Prevalence of thinness among children ages 10 to 19")
plot_log(training$thinness..1.19.years)+ 
  xlab("Prevalence of thinness among children ages 10 to 19")

# Thinness 5-9 Variable

plot_original(training$thinness.5.9.years)+ 
  xlab("Prevalence of thinness among children ages 5 to 9")
plot_original_interaction(training$thinness.5.9.years)+ 
  xlab("Prevalence of thinness among children ages 5 to 9")
plot_poly(training$thinness.5.9.years)+ 
  xlab("Prevalence of thinness among children ages 5 to 9")
plot_log(training$thinness.5.9.years)+ 
  xlab("Prevalence of thinness among children ages 5 to 9")

# Income Composition of Resources Variable

plot_original(training$Income.composition.of.resources) + 
  xlab("Human Development Index in terms of income composition of resources")
plot_original_interaction(training$Income.composition.of.resources) + 
  xlab("Human Development Index in terms of income composition of resources")
plot_poly(training$Income.composition.of.resources)+ 
  xlab("Human Development Index in terms of income composition of resources")
plot_log(training$Income.composition.of.resources)+ 
  xlab("Human Development Index in terms of income composition of resources")

# Schooling Variable

plot_original(training$Schooling) + xlab("Years in School")
plot_original_interaction(training$Schooling) + xlab("Years in School")
plot_poly(training$Schooling)+ xlab("Years in School")
plot_log(training$Schooling)+ xlab("Years in School")
```

Original Model for comparison
```{r Original Model}
original.model <- lm(Life.expectancy~ Polio + Alcohol + BMI +
                       Diphtheria+ Schooling + HIV.AIDS + GDP +
                       thinness..1.19.years + Measles + Total.expenditure + 
                       Income.composition.of.resources, training)

plot(original.model)
```
Fitting a complex model with polynomials up to 2 & logged variables

```{r Complex poly 2 log Model}
poly2logmodel <- lm(Life.expectancy~ poly(Polio,2)+
                      poly(Alcohol,2)+
                      poly(BMI,2)+
                      poly(Diphtheria,2)+
                      poly(Schooling,2)+
                      log(HIV.AIDS) +
                      log(GDP) +
                      log(thinness..1.19.years) +
                      Measles +
                      Total.expenditure+
                      Income.composition.of.resources, training)

plot(poly2logmodel)

```

Fitting a complex model with polynomials up to 7 & logged variables

```{r Complex poly 7 log Model}
poly7logmodel <- lm(Life.expectancy~ poly(Polio,7)+
                      sqrt(Alcohol)+
                      poly(BMI,7)+
                      poly(Diphtheria,5)+
                      poly(Schooling,7)+
                      log(HIV.AIDS) +
                      log(GDP) +
                      log(thinness..1.19.years) +
                      Measles +
                      Total.expenditure+
                      Income.composition.of.resources, training)

plot(poly7logmodel)

```

Fitting interaction terms on regular model
```{r Original with interaction Model}
interaction.model <- lm(Life.expectancy~ Polio:Status+Alcohol:Status+BMI:Status+
                          Diphtheria:Status+Schooling:Status+
                          HIV.AIDS:Status+GDP:Status+
                          thinness..1.19.years:Status+
                          Measles:Status+Total.expenditure:Status+
                          Income.composition.of.resources:Status, training)

plot(interaction.model)

```
Fitting interaction on poly2logmodel
```{r Complex poly 2 log with interaction Model}
poly2log.interaction.model <- lm(Life.expectancy~ poly(Polio,2):Status+
                      poly(Alcohol,2):Status+
                      poly(BMI,2):Status+
                      poly(Diphtheria,2):Status+
                      poly(Schooling,2):Status+
                      log(HIV.AIDS):Status +
                      log(GDP):Status +
                      log(thinness..1.19.years):Status +
                      Measles:Status +
                      Total.expenditure:Status+
                      Income.composition.of.resources:Status, training)

plot(poly2log.interaction.model)

```
Fitting interaction on poly7logmodel
```{r Complex poly 7 log with interaction Model}
poly7log.interaction.model <- lm(Life.expectancy~ poly(Polio,7):Status+
                      sqrt(Alcohol):Status+
                      poly(BMI,7):Status+
                      poly(Diphtheria,5):Status+
                      poly(Schooling,7):Status+
                      log(HIV.AIDS):Status +
                      log(GDP):Status +
                      log(thinness..1.19.years):Status +
                      Measles:Status +
                      Total.expenditure:Status+
                      Income.composition.of.resources:Status, training)

plot(poly7log.interaction.model)

```

Fitting a complex model with polynomials up to 7 & logged variables and categorical variable Status

```{r Complex poly 7 add Status log Model}
status.poly7logmodel <- lm(Life.expectancy~ Status + poly(Polio,7)+
                      sqrt(Alcohol)+
                      poly(BMI,7)+
                      poly(Diphtheria,5)+
                      poly(Schooling,7)+
                      log(HIV.AIDS) +
                      log(GDP) +
                      log(thinness..1.19.years) +
                      Measles +
                      Total.expenditure+
                      Income.composition.of.resources, training)

plot(status.poly7logmodel)

```
Simple Model
```{r}
simple <- lm(Life.expectancy ~ HIV.AIDS + Alcohol:Status + Total.expenditure, training)
plot(simple)
```
Predictions
```{r Predictions}
# RMSE of Full Model
full.model_Preds = predict(full.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = full.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Penalized Regression Model
glmnet.fit.model_Preds = predict(glmnet.fit.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = glmnet.fit.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Forward Selection Model
fwd.select.model_Preds = predict(fwd.select.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = fwd.select.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Backward Selection Model
bck.select.model_Preds = predict(bck.select.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = bck.select.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Stepwise Selection Model
sw.select.model_Preds = predict(sw.select.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = sw.select.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Original Model
original.model_Preds = predict(original.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = original.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Poly2Log Model
poly2logmodel_Preds = predict(poly2logmodel, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = poly2logmodel_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Poly7Log Model

poly7logmodel_Preds = predict(poly7logmodel, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = poly7logmodel_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Original with Interaction Model
interaction.model_Preds = predict(interaction.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = interaction.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Poly2Log Model with interactions
poly2log.interaction.model_Preds = predict(poly2log.interaction.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = poly2log.interaction.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))


# RMSE of Poly7Log Model with interactions
poly7log.interaction.model_Preds = predict(poly7log.interaction.model, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = poly7log.interaction.model_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))

# RMSE of Poly7Log Model with Categorical variable Status added (not as interaction)
status.poly7logmodel_Preds = predict(status.poly7logmodel, newdata = validate)

MSPE = data.frame(Observed = validate$Life.expectancy, 
                  Predicted = status.poly7logmodel_Preds)
MSPE$Residual = MSPE$Observed - MSPE$Predicted
MSPE$SquaredResidual = MSPE$Residual^2
sqrt(mean(MSPE$SquaredResidual))
```
#knn with training 
```{r}
library(caret)
fit_cont1 = trainControl(method = "repeatedcv", number = 10, repeats = 1)

set.seed(1364)

knnfit1 = train(Life.expectancy~Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling + logGDP, data =training, method = "knn", trControl = fit_cont1, tuneGrid = expand.grid(k = c(1:30)))

plot(knnfit1)

updateval = validate[,c("Life.expectancy", "Adult.Mortality", "infant.deaths", "Alcohol", "percentage.expenditure", "Measles", "BMI", "under.five.deaths", "Polio", "Total.expenditure", "Diphtheria", "HIV.AIDS", "thinness..1.19.years","thinness.5.9.years",  "Income.composition.of.resources", "Schooling", "logGDP")]

prediction = predict(knnfit1, newdata = updateval)

MSPE = data.frame(Observed = validate$Life.expectancy, Predicted = prediction)

MSPE$Residual = MSPE$Observed - MSPE$Predicted

MSPE$SquaredResidual = MSPE$Residual^2

sqrt(mean(MSPE$SquaredResidual))

#prediction = predict(knnfit1, newdata = validate)
#cf = confusionMatrix(prediction, updateval$Life.expectancy)
#cf
#print(cf)

```