---
title: "Proj1"
author: "Thibeaux"
date: "2023-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(GGally)
library(boot)
library(ggplot2)
library(tidyverse)
library(lmboot)
library(caret)
library(naniar)
library(utils)
library(stats)
library(corrplot)
library(ISLR)
library(car)
library(olsrr)
```

##Reading in Dataset
```{r}
life = read.csv('https://github.com/athibeaux/MSDS-DS6372/raw/main/Life_Expectancy.csv')
```
```{r}
ggplot(data = life) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))
```
##Upon looking at the graph of the original data set, it appears that there needs to be a log transformation on the X or the GDP as we are interested in seeing the relation between Life Expenctancy and GDP. 

##Checking Data Types
```{r}
str(life)
vis_miss(life)
dim(life)
```

#Imputing using Median
```{r}
#GDP 15% [17]  keep GDP to have it Imputed even if quite high percentage, assuming it is crucial to predicting Life.expectancy as richer countries have better health access/Medicine and tech. The numbers appear to be GDP per capita which helps as it addresses GDP/Population. GDP per Capita and Population would be too closely related and prob attribute to covariance.  

#Adjusting text angle to vis_miss
imputeMedian= preProcess(life[,-c(1:4,9)],method="medianImpute") #predictors 1:3, 9, 17 and response is 4
cleandataMedian = predict(imputeMedian,newdata=life)
dim(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#Literature says that over 10% missing data can contribute to bias 
#HepatitsB [9] at 19% , Population 22% [18]. 
#Removing columns 9 and 18
cleandataMedian = cleandataMedian[,-c(18,9)]
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#removing last NA
cleandataMedian = na.omit(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

```

Before and after log transforming GDP, with cleandataMedian:
```{r}
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))

#Log transformation on GDP 
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = log(GDP), y = Life.expectancy))

```

#Imputing and Removing
```{r}
#Imputing all save for columns 1:4 and Removing last Na in 
imputeMedian= preProcess(life[,-c(1:4)],method="medianImpute") #predictors 1:4 and response is 4
cleandataMedian1 = predict(imputeMedian,newdata=life)
dim(cleandataMedian1)
vis_miss(cleandataMedian1) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#removing last NA <0,1% 
cleandataMedian1 = na.omit(cleandataMedian1)
vis_miss(cleandataMedian1) + theme(axis.text.x = element_text(angle = 90, hjust = 0))
dim(cleandataMedian1)

```


##Correlation Matrix:
```{r}
cor <- cor(cleandataMedian[,c(4,5:20)])
corrplot(cor, method = "square",  tl.srt = 50, tl.col = "black", tl.cex = 0.6, title = "Correlation of Variables", mar=c(0,0,1,0))
```
GGPairs:
```{r}
#commented out for knitting
#ggpairs(cleandataMedian[,4:20])
```
Create categorical variable from year:
```{r}

summary(cleandataMedian$Year)

#Labeled the breaks in years into 2 year buckets
cleandataMedian$DualYears <- cut(as.numeric(cleandataMedian$Year), breaks=c(0, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015), labels=c('2000-2001', '2002-2003', '2004-2005', '2006-2007', '2008-2009', '2010-2011','2012-2013','2014-2015'))

cleandataMedian %>% ggplot() + geom_bar(aes(x = DualYears))

summary(cleandataMedian$DualYears)

```
Function for Individual T-Tests on each variable
```{r}
# Create the Function
tstt <- function(var1) {
  return(t.test(var1~Life.expectancy, data = cleandataMedian,conf.level = .95,var.equal=FALSE))
}

```
Fitting a preliminary model of non-redundant variables
```{r prelim model}
model <- lm(Life.expectancy~Status + Alcohol + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling, cleandataMedian)

summary(model)
plot(model)
cleandataMedian %>% ggplot(aes(Life.expectancy, Polio)) + geom_smooth() + geom_point()

vif(model)
```
Feature Selection Tools: Penalized Regression
```{r Penalized Regression}

# Penalized Regression

#Setting kfold parameters
fitControl<-trainControl(method="repeatedcv",number=5,repeats=1) 

#Fitting glmnet
set.seed(1234)
glmnet.fit<-train(log(Life.expectancy)~Adult.Mortality + Alcohol + Measles + BMI + under.five.deaths + Polio + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,
                  data=cleandataMedian,
                  method="glmnet",
                  trControl=fitControl
)

glmnet.fit
plot(glmnet.fit)


#Investigating coefficients
opt.pen<-glmnet.fit$finalModel$lambdaOpt #penalty term
coef(glmnet.fit$finalModel,opt.pen)


summary(lm(log(Life.expectancy)~Alcohol + Measles + BMI + under.five.deaths + Polio + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,
          data=cleandataMedian))

#Lets force a LASSO model and add complexity
set.seed(1234)
glmnet.fit2<-train(log(Life.expectancy)~Alcohol + Measles + BMI + under.five.deaths:Status + poly(Polio,2) + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,
                  data=cleandataMedian,
                  method="glmnet",
                  trControl=fitControl,
                  tuneGrid=expand.grid(data.frame(alpha=1,lambda=seq(0,.05,.001)))
)

glmnet.fit2
plot(glmnet.fit2)

opt.pen<-glmnet.fit2$finalModel$lambdaOpt   #penalty term
coef(glmnet.fit2$finalModel,opt.pen)

# Different way to do GLMNET

x=model.matrix(Life.expectancy~Alcohol + Measles + BMI + under.five.deaths:Status + poly(Polio,2) + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,cleandataMedian)[,-1]
y=log(cleandataMedian$Life.expectancy)

library(glmnet)
set.seed(1234)

#grid=10^seq(10,-2, length =100)
#lasso.mod=glmnet(x,y,alpha=1, lambda =grid)

cv.out=cv.glmnet(x,y,alpha=1)
plot(cv.out)
bestlambda<-cv.out$lambda.1se
coef(cv.out,s=bestlambda)
```
Feature Selection Tools: Forward Selection
```{r Forward Selection}
library(leaps)
reg.fwd=regsubsets(log(Life.expectancy)~Alcohol + Measles + BMI + under.five.deaths + Polio + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,
                   data=cleandataMedian,method="forward",nvmax=12)

summary(reg.fwd)$adjr2
summary(reg.fwd)$rss
summary(reg.fwd)$bic


par(mfrow=c(1,3))
bics<-summary(reg.fwd)$bic
plot(1:12,bics,type="l",ylab="BIC",xlab="# of predictors")
index<-which(bics==min(bics))
points(index,bics[index],col="red",pch=10)

adjr2<-summary(reg.fwd)$adjr2
plot(1:12,adjr2,type="l",ylab="Adjusted R-squared",xlab="# of predictors")
index<-which(adjr2==max(adjr2))
points(index,adjr2[index],col="red",pch=10)

rss<-summary(reg.fwd)$rss
plot(1:12,rss,type="l",ylab="train RSS",xlab="# of predictors")
index<-which(rss==min(rss))
points(index,rss[index],col="red",pch=10)
```

Splitting into Train/Test data:
```{r Train Test Data}
set.seed(1234)
trainIndex<-createDataPartition(cleandataMedian$Life.expectancy,p=.5,list=F)  #p: proportion of data in train

training<-cleandataMedian[trainIndex,]
validate<-cleandataMedian[-trainIndex,]

fwd.train=regsubsets(log(Life.expectancy)~Alcohol + Measles + BMI + under.five.deaths + Polio + Total.expenditure +
                    Diphtheria + HIV.AIDS + GDP + thinness..1.19.years + Income.composition.of.resources + Schooling,
                    data=training,method="forward",nvmax=12)

#Creating a prediction function 
predict.regsubsets =function (object , newdata ,id ,...){
  form=as.formula (object$call [[2]])
  mat=model.matrix(form ,newdata )
  coefi=coef(object ,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}

valMSE<-c()
#note my index, i, is to 12 since that is how many predictors I went up to during fwd selection
for (i in 1:12){
  predictions<-predict.regsubsets(object=fwd.train,newdata=validate,id=i) 
  valMSE[i]<-mean((log(validate$Life.expectancy)-predictions)^2)
}
par(mfrow=c(1,1))
plot(1:12,sqrt(valMSE),type="l",xlab="# of predictors",ylab="test vs train RMSE",ylim=c(0.3,.9))
index<-which(valMSE==min(valMSE))
points(index,sqrt(valMSE[index]),col="red",pch=10)

trainMSE<-summary(fwd.train)$rss/nrow(training)
lines(1:12,sqrt(trainMSE),lty=3,col="blue")  
```
Adding Complexity to the Model
```{r Transformation Exploration}
# Transformations using Polio as a test

transformations <- cleandataMedian

# Transforming with Polynomial using a ^2
transformations$polio2 <- transformations$Polio^2
transformations %>% ggplot(aes(x=polio2, y=Life.expectancy)) +
geom_point() + geom_smooth()

# Transforming using Log
transformations$logPolio <- log(transformations$Polio)

transformations %>% ggplot(aes(x=logPolio, y=Life.expectancy)) +
geom_point() + geom_smooth()

# Testing if it's the same as logging inside the plot
transformations %>% ggplot(aes(x=log(Polio), y=Life.expectancy)) +
geom_point() + geom_smooth()
```
Creating plot functions
```{r Creating Plot Functions}

# Original Data plot
plot_original <- function(var1) {
  return(cleandataMedian %>% ggplot(aes(x=var1, 
                                        y = Life.expectancy, 
                                        color=Status)) + 
           geom_point() + geom_smooth ()
)
}

plot_original(transformations$Polio)

# Creating function to plot variable as polynomials 

plot_poly <- function(var1) {
  return(cleandataMedian %>% ggplot(aes(x=var1^2, 
                                      y = Life.expectancy,
                                      color=Status)) + 
         geom_point() + geom_smooth ()
)
}

plot_poly(transformations$Polio)

# Creating function to plot variables as logged
plot_log <- function(var1) {
  return(cleandataMedian %>% ggplot(aes(x=log(var1), 
                                      y = Life.expectancy, 
                                      color=Status)) + 
         geom_point() + geom_smooth ()
)
}

plot_log(transformations$Polio)
```
Putting each Variable into the Function
```{r Variables into Function}
# Infant Deaths Variable

plot_original(cleandataMedian$infant.deaths)
plot_poly(cleandataMedian$infant.deaths)
plot_log(cleandataMedian$infant.deaths)

# Alcohol Variable

plot_original(cleandataMedian$Alcohol)
plot_poly(cleandataMedian$Alcohol)
plot_log(cleandataMedian$Alcohol)

# Percentage of Expenditures Variable

plot_original(cleandataMedian$percentage.expenditure)
plot_poly(cleandataMedian$percentage.expenditure)
plot_log(cleandataMedian$percentage.expenditure)

# Measles Variable

plot_original(cleandataMedian$Measles)
plot_poly(cleandataMedian$Measles)
plot_log(cleandataMedian$Measles)

# BMI Variable

plot_original(cleandataMedian$BMI)
plot_poly(cleandataMedian$BMI)
plot_log(cleandataMedian$BMI)

# Under Five Deaths Variable

plot_original(cleandataMedian$under.five.deaths)
plot_poly(cleandataMedian$under.five.deaths)
plot_log(cleandataMedian$under.five.deaths)

# Total Expenditures Variable

plot_original(cleandataMedian$Total.expenditure)
plot_poly(cleandataMedian$Total.expenditure)
plot_log(cleandataMedian$Total.expenditure)

# Diphtheria Variable

plot_original(cleandataMedian$Diphtheria)
plot_poly(cleandataMedian$Diphtheria)
plot_log(cleandataMedian$Diphtheria)

# HIV AIDS Variable

plot_original(cleandataMedian$HIV.AIDS)
plot_poly(cleandataMedian$HIV.AIDS)
plot_log(cleandataMedian$HIV.AIDS)

# GDP Variable

plot_original(cleandataMedian$GDP)
plot_poly(cleandataMedian$GDP)
plot_log(cleandataMedian$GDP)

# Thinness 10-19 Years Variable

plot_original(cleandataMedian$thinness..1.19.years)
plot_poly(cleandataMedian$thinness..1.19.years)
plot_log(cleandataMedian$thinness..1.19.years)

# Thinness 5-9 Variable

plot_original(cleandataMedian$thinness.5.9.years)
plot_poly(cleandataMedian$thinness.5.9.years)
plot_log(cleandataMedian$thinness.5.9.years)

# Income Composition of Resources Variable

plot_original(cleandataMedian$Income.composition.of.resources)
plot_poly(cleandataMedian$Income.composition.of.resources)
plot_log(cleandataMedian$Income.composition.of.resources)

# Schooling Variable

plot_original(cleandataMedian$Schooling)
plot_poly(cleandataMedian$Schooling)
plot_log(cleandataMedian$Schooling)
```

Fitting a complex model with polynomials, logged variables, and interaction

```{r Complex Model}
poly2logmodel <- lm(Life.expectancy~ poly(Polio,2)+
                      poly(Alcohol,2)+
                      poly(BMI,2)+
                      poly(Diphtheria,2)+
                      poly(Income.composition.of.resources, 2)+
                      poly(Schooling,2)+
                      log(HIV.AIDS) +
                      log(GDP) +
                      log(thinness..1.19.years) +
                      Measles +
                      Total.expenditure, cleandataMedian)

plot(poly2logmodel)
```
