---
title: "Project 1"
author: "Xavier Mojica"
date: "2023-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(GGally)
library(boot)
library(ggplot2)
library(tidyverse)
library(lmboot)
library(lattice)
library(caret)
library(naniar)
library(utils)
library(stats)
```

##Reading in Dataset
```{r}
setwd("/Users/xaviermojica/Desktop/Stats2/Project1") #/Users/xaviermojica/
life = read.csv("Life Expectancy Data (1).csv")
```


```{r}
ggplot(data = life) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))
```
##Upon looking at the graph of the original data set, it appears that there needs to be a log transformation on the X or the GDP as we are interested in seeing the relation between Life Expenctancy and GDP. 

##Checking Data Types
```{r}
str(life)
vis_miss(life)
dim(life)
View(life)
```

#Imputing using Median
```{r}
#GDP 15% [17]  keep GDP to have it Imputed even if quite high percentage, assuming it is crucial to predicting Life.expectancy as richer countries have better health access/Medicine and tech. The numbers appear to be GDP per capita which helps as it addresses GDP/Population. GDP per Capita and Population would be too closely related and prob attribute to covariance.  
#Adjusting text angle to vis_miss
imputeMedian= preProcess(life[,-c(1:4,9)],method="medianImpute") #predictors 1:4, 9 and response is 4
cleandataMedian = predict(imputeMedian,newdata=life)
dim(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#Literature says that over 10% missing data can contribute to bias 
#HepatitsB [9] at 19% , Population 22% [18]. 
#Removing columns 9 and 18
cleandataMedian = cleandataMedian[,-c(18,9)]
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

#removing last NA
cleandataMedian = na.omit(cleandataMedian)
vis_miss(cleandataMedian) + theme(axis.text.x = element_text(angle = 90, hjust = 0))

```


```{r}
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = GDP, y = Life.expectancy))

#Converting GDP to Log 
cleandataMedian$logGDP = log(cleandataMedian$GDP)
#converting Life.expectancy to log 
cleandataMedian$logLife.expectancy = log(cleandataMedian$Life.expectancy)

#Log transformation on GDP 
ggplot(data = cleandataMedian) + geom_point(mapping = aes(x = logGDP, y = logLife.expectancy))




```


#Split Data
```{r}
set.seed(1234)
trainIndex<-createDataPartition(cleandataMedian$Life.expectancy,p=.8,list=F)

training<-cleandataMedian[trainIndex,]
validate<-cleandataMedian[-trainIndex,]

```
#Multivariable Plots on Full Data
```{r}
library(ISLR)
library(data.table)
library(GGally)
library(ggplot2)

ggpairs(training[,4:8], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,9:12)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,13:16)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))

ggpairs(training[,c(4,17:20)], lower = list(continuous = wrap("points", color = "red", alpha = 0.5), combo = wrap("box", color = "orange", alpha = 0.3), discrete = wrap("facetbar", color = "yellow", alpha = 0.3) ),diag = list(continuous = wrap("densityDiag",  color = "blue", alpha = 0.5)))
#ggpairs(cleandataMedian[,5:22], upper = list(continuous = wrap("cor", size = 4.75, align_percent = 1)))
#ggscatmat(cleandataMedian, columns = 4:10)
```



#Residuals for chosen MLR model 
```{r}
set.seed(2345)
eightVar = lm(Life.expectancy~HIV.AIDS+Schooling+Alcohol+BMI+Polio+Diphtheria+logGDP+thinness..1.19.years+Income.composition.of.resources, data = training)
summary(eightVar)
confint(eightVar)

#Visuals for Residuals
plot(eightVar)

```

#Bests MLR 
```{r}
set.seed(2323)
lessvar = lm(Life.expectancy~HIV.AIDS+Schooling+BMI+Diphtheria+logGDP+thinness..1.19.years+Income.composition.of.resources, data = training)
summary(lessvar)
confint(lessvar)

#Visuals for Residuals
plot(lessvar)

```






#Forward, Backward, Stepwise Selection 
```{r}
library(MASS)
#Full Model 
set.seed(1246)
fitFull = lm(Life.expectancy~Adult.Mortality + 
     infant.deaths + Alcohol + percentage.expenditure + Measles + 
     BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling + logGDP,training)

stepup = stepAIC(fitFull, direction = "forward", steps = 2000)

stepdown= stepAIC(fitFull, direction = "backward", steps = 2000)

stepboth = stepAIC(fitFull, direction = "both", steps = 2000)

#summary for each model 
up = summary(stepup)
up
down = summary(stepdown)
down
both = summary(stepboth)
both
```

```{r}
fitFull

olsrr::ols_step_forward_aic(fitFull)
```

#KNN 
```{r}
#library(caret)

#fit_cont = trainControl(method = "repeatedcv", number = 10, repeats = 1)
#set.seed(136)

#knnfit = train(Life.expectancy~Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling + logGDP, data =cleandataMedian, method = "knn", trControl = fit_cont, tuneGrid = expand.grid(k = c(1:30)))

#knnfit

#plot(knnfit)
```

#knn with training 
```{r}
library(caret)
fit_cont1 = trainControl(method = "repeatedcv", number = 10, repeats = 1)

set.seed(1364)

knnfit1 = train(Life.expectancy~Adult.Mortality + infant.deaths + Alcohol + percentage.expenditure + Measles + BMI + under.five.deaths + Polio + Total.expenditure + Diphtheria + HIV.AIDS + thinness..1.19.years + thinness.5.9.years + Income.composition.of.resources + Schooling + logGDP, data =training, method = "knn", trControl = fit_cont1, tuneGrid = expand.grid(k = c(1:30)))


plot(knnfit1)


updateval = validate[,c("Life.expectancy", "Adult.Mortality", "infant.deaths", "Alcohol", "percentage.expenditure", "Measles", "BMI", "under.five.deaths", "Polio", "Total.expenditure", "Diphtheria", "HIV.AIDS", "thinness..1.19.years","thinness.5.9.years",  "Income.composition.of.resources", "Schooling", "logGDP")]

prediction = predict(knnfit1, newdata = validate)
#cf = confusionMatrix(prediction, updateval$Life.expectancy)
#cf
#print(cf)

```
```

#World Map, Color plotting 
```{r}
#Creating the World 
#library(ggplot2)
#library(tidyverse)
#library(ggthemes)

#world_map = map_data("world") %>% filter(! long > 180)

#countries = world_map %>% distinct(region) %>% rowid_to_column()

#countries %>% ggplot(aes(fill = rowid, map_id = region)) + geom_map(map = world_map) + expand_limits(x = world_map$long, y = world_map$lat) + coord_map("moll") +theme_map()


```


#Color world plotting with full data
```{r}
library(ggplot2)
library(tidyverse)

#rename training for the map
dataforcolmap = training
#but first renaming column Country in dataforcolmap 
colnames(dataforcolmap)[1] = "region"
view(dataforcolmap)


#getting map data for plotting 
mapdata = map_data("world")
view(mapdata)

#joining map data with dataforcolmap 
mapdata = left_join(mapdata,dataforcolmap, by = "region")
view(mapdata)

#filtering out NAs for life expectancy , status, Income.composition.of.resources
#Life Exp
mapdata1 = mapdata %>% filter(!is.na(mapdata$Life.expectancy))

#Status 
mapdata2 = mapdata %>% filter(!is.na(mapdata$Status))

#Income
mapdata3 = mapdata %>% filter(!is.na(mapdata$Income.composition.of.resources))


#mapping mapdata1 for Life Exp 

map1 = ggplot(mapdata1, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Life.expectancy), color = "black")+ theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Life Expectancy per Country") + scale_fill_gradient(low = "red", high = "yellow") + guides(fill=guide_legend(title="Life Expectancy"))
map1



#mapping mapdata2 for Status 

mapStatus = ggplot(mapdata2, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Status, col = "orange"), color = "black") + theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Country's Status: Developed v. Developing") 
mapStatus


#mapping mapdata3 for Income Composition of Resources 

mapIncome = ggplot(mapdata3, aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = Income.composition.of.resources), color = "black")+ theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), axis.title.y = element_blank(), axis.title.x = element_blank()) + ggtitle("Income Composition of Resources per Country") + scale_fill_gradient(low = "red", high = "yellow")+ guides(fill=guide_legend(title="Income Composition of Resources"))
mapIncome

```

```{}


```




```{r}
# Using rpart library
#treeimb <- rpart(ExplVar ~ ., data = train)
#pred.treeimb <- predict(treeimb, newdata = test)

```






